Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-06-26 pirouette_article.tex
Find bracket error

BEAST [\cite{drummond2007beast}] is a typical Bayesian phylogenetics tool, 
that needs both character data and priors to infer 
a posterior distribution of phylogenies.
Specifically, for the species tree prior - which describes 
the process of diversification - 
BEAST has built-in priors such as the Yule [\cite{yule}] and 
(constant-rate) birth-death [\cite{nee1994reconstructed}] models
\new{as well as coalescent priors.}
These simple tree priors are among the most commonly used, 
as they represent some biologically realistic processes (e.g. 
viewing diversification as a branching process), while being computationally fast.

To allow users to extend the functionalities of BEAST
using plug-ins, BEAST2 was written [\cite{bouckaert2019beast}]
(with BEAST and BEAST2 still independently being developed further).
For example, one can add novel diversification models 
by writing a BEAST2 plugin that contains the likelihood formula of a phylogeny under the novel diversification model, 
i.e. the prior probability of a species tree.
Plugins have been provided, for instance, for the calibrated Yule model [\cite{heled2015calibrated}],
  BD with incomplete sampling [\cite{stadler2009incomplete}],
  BD serial sampling [\cite{stadler2012estimating}],
  BD serial skyline [\cite{stadler2013birth}]
  Fossilized BD process [\cite{gavryushkina2014bayesian}], and
  BD SIR [\cite{kuhnert2014simultaneous}].
Many other diversification models (and their associated likelihood algorithms) 
have been developed, e.g., models in which diversification is time-dependent [\cite{nee1994reconstructed,rabosky2008explosive}],
or diversity-dependent [\cite{etienne2012diversity}],
or where diversification rates change for specific lineages 
and their descendants [\cite{etienne2012conceptual, rabosky2014automatic, alfaro2009nine, laudanno2020sls}].
Other models treat speciation as a process that takes 
time [\cite{rosindell2010protracted, etienne2012prolonging, lambert2015reconstructed}],
or where diversification rates
depends on one or more traits [\cite{maddison2007estimating, fitzjohn2012diversitree, herrera2018detecting} ].
These are, however, not yet available as tree priors in BEAST2, for reasons explained below. In this paper, we present methodology to determine whether such new plug-ins are needed, or whether currently available plug-ins are sufficient. We show this using the Yule and BD species tree priors, but our methods can be used with other built-in tree priors as well.

@@

\giovanni{Do we need the library calls in the listing?}
\begin{lstlisting}[language=R]
library(ggplot2)
library(pirouette)

@@

# Plot
pir_plot(pir_out)

# Save
pir_save(
  phylogeny = phylogeny,
  pir_params = pir_params,
  pir_out = pir_out,
  folder_name = "example_30"
)
\end{lstlisting}

The DD tree generated by this code is shown in Figure \ref{fig:dd_tree}. \iffalse % wordcut (this info is already in listing)
,which has an arbitrarily chosen crown age of 10 time units and 6 tips 
for an extinction rate of 0.1. The carrying-capacity is also set at 6. The initial speciation rate $\lambda_0$ is chosen such that the expected number of species in a constant-rate BD model would be equal to the number of tips, which amounts to $\lambda_0 = 0.63$.
\fi

@@

Interpreting the results of \verb;pirouette; is up to the user; 
\verb;pirouette; does not answer the question 
whether the inference error is too large to trust the inferred tree. The user is encouraged to use different statistics to measure the error. The nLTT statistic is
a promising starting point, as it can compare any two trees and 
results in an error distribution of known range, but one may also explore other statistics.
In principle, \verb;pirouette; allows for this, but in our example we used a diversification model (DD) that only deviates from the Yule and BD models in the temporal branching pattern, not in the topology.
\new{
  For models that make different predictions on topology, 
  the twinning process should be modified accordingly.
}
\new{
As noted in the introduction, Duch\^{e}ne and 
colleagues [\cite{duchene2018phylodynamic}]
also developed a method to assess the adequacy of a tree model
on empirical trees. They simulate trees from the posterior distribution of 
the parameters and then compare this to the originally inferred tree using 
tree statistics, to determine whether the assumed tree model in inference 
indeed generates the tree as inferred. This is useful if these trees match, 
but when they do not, this does not mean that the inferred tree is incorrect; 
if sufficient data is available the species tree prior may not be important, 
and hence the inference may be adequate 
even though the assumed species tree prior is not. 
In short, the approach is applied to empirical trees and 
compares the posterior and prior distribution of trees (with the latter 
generated with the posterior parameters!).
\verb;pirouette; aims to identify when assuming standard priors for the 
species tree leads to incorrect inference if one believes more complex 
diversification models are operating than can be currently accommodated in 
inference. In short, our approach applies to simulated trees and compares 
the posterior distributions of trees generated with 
a standard and non-standard model, but inferred with a standard one. 
The two methods therefore complement one another.
}
\giovanni{I don't understand this sentence: "pirouette aims to identify when assuming standard priors for the 
species tree leads to incorrect inference if one believes more complex 
diversification models are operating than can be currently accommodated in 
inference". Maybe shortening it, it can become clearer}
However, we note that the \verb;pirouette; pipeline is not restricted 
to exploring the effects of a new species tree model. 
The pipeline can also be used to explore the effects of non-standard 
clock or site models, such as relaxed clock models with a non-standard 
distribution, correlated substitutions on sister lineages, or elevated 
substitutions rates during speciation events. 
It is, however, beyond the scope of this paper to discuss all these options 
in more detail.

>>>>

	BEAST [\cite{drummond2007beast}] is a typical Bayesian phylogenetics tool, 
	that needs both character data and priors to infer 
	a posterior distribution of phylogenies.
	Specifically, for the species tree prior - which describes 
	the process of diversification - 
	BEAST has built-in priors such as the Yule [\cite{yule}] and 
	(constant-rate) birth-death [\cite{nee1994reconstructed}] models
	as well as coalescent priors.
	These simple tree priors are among the most commonly used, 
	as they represent some biologically realistic processes (e.g. 
	viewing diversification as a branching process), while being computationally fast.

	To allow users to extend the functionalities of BEAST
	using plug-ins, BEAST2 was written [\cite{bouckaert2019beast}]
	(with BEAST and BEAST2 still independently being developed further).
	For example, one can add novel diversification models 
	by writing a BEAST2 plugin that contains the likelihood formula of a phylogeny under the novel diversification model, 
	i.e. the prior probability of a species tree.
	Plugins have been provided, for instance, for the calibrated Yule model [\cite{heled2015calibrated}],
	  BD with incomplete sampling [\cite{stadler2009incomplete}],
	  BD serial sampling [\cite{stadler2012estimating}],
	  BD serial skyline [\cite{stadler2013birth}]
	  Fossilized BD process [\cite{gavryushkina2014bayesian}], and
	  BD SIR [\cite{kuhnert2014simultaneous}].
	Many other diversification models (and their associated likelihood algorithms) 
	have been developed, e.g., models in which diversification is time-dependent [\cite{nee1994reconstructed,rabosky2008explosive}],
	or diversity-dependent [\cite{etienne2012diversity}],
	or where diversification rates change for specific lineages 
	and their descendants [\cite{etienne2012conceptual, rabosky2014automatic, alfaro2009nine, laudanno2020sls}].
	Other models treat speciation as a process that takes 
	time [\cite{rosindell2010protracted, etienne2012prolonging, lambert2015reconstructed}],
	or where diversification rates
	depends on one or more traits [\cite{maddison2007estimating, fitzjohn2012diversitree, herrera2018detecting}].
	These are, however, not yet available as tree priors in BEAST2, 
	for reasons explained below. In this paper, 
	we present methodology to determine whether such new plug-ins are needed, 
	or whether currently available plug-ins are sufficient. 
	We show this using the Yule and BD species tree priors, 
	but our methods can be used with other built-in tree priors as well.

@@

\giovanni{
  Do we need the library calls in the listing?
}
\richel{
  I simplified the code and only call pirouette now. 
  I think this is cleaner then using a scope operator.
}
\begin{lstlisting}[language=R]
library(pirouette)

@@

# Plot
pir_plot(pir_out)
\end{lstlisting}

The DD tree generated by this code is shown in Figure \ref{fig:dd_tree}.

@@

Interpreting the results of \verb;pirouette; is up to the user; 
\verb;pirouette; does not answer the question 
whether the inference error is too large to trust the inferred tree. The user is encouraged to use different statistics to measure the error. The nLTT statistic is
a promising starting point, as it can compare any two trees and 
results in an error distribution of known range, but one may also explore other statistics.
In principle, \verb;pirouette; allows for this, but in our example we used a diversification model (DD) that only deviates from the Yule and BD models in the temporal branching pattern, not in the topology.
\new{
  For models that make different predictions on topology, 
  the twinning process should be modified accordingly.
}
\new{
	As noted in the introduction, Duch\^{e}ne and 
	colleagues [\cite{duchene2018phylodynamic}]
	also developed a method to assess the adequacy of a tree model
	on empirical trees. They simulate trees from the posterior distribution of 
	the parameters and then compare this to the originally inferred tree using 
	tree statistics, to determine whether the assumed tree model in inference 
	indeed generates the tree as inferred. This is useful if these trees match, 
	but when they do not, this does not mean that the inferred tree is incorrect; 
	if sufficient data is available the species tree prior may not be important, 
	and hence the inference may be adequate 
	even though the assumed species tree prior is not. 
	In short, the approach is applied to empirical trees and 
	compares the posterior and prior distribution of trees (with the latter 
	generated with the posterior parameters!).
	\verb;pirouette; aims to identify when assuming standard priors for the 
	species tree leads to incorrect inference if one believes more complex 
	diversification models are operating than can be currently accommodated in 
	inference. In short, our approach applies to simulated trees and compares 
	the posterior distributions of trees generated with 
	a standard and non-standard model, but inferred with a standard one. 
	The two methods therefore complement one another.
}
\giovanni{
	I don't understand this sentence: "pirouette aims to identify when assuming standard priors for the 
	species tree leads to incorrect inference if one believes more complex 
	diversification models are operating than can be currently accommodated in 
	inference". Maybe shortening it, it can become clearer
}
However, we note that the \verb;pirouette; pipeline is not restricted 
to exploring the effects of a new species tree model. 
The pipeline can also be used to explore the effects of non-standard 
clock or site models, such as relaxed clock models with a non-standard 
distribution, correlated substitutions on sister lineages, or elevated 
substitutions rates during speciation events. 
It is, however, beyond the scope of this paper to discuss all these options 
in more detail.
